Disease Knowledge Transfer across Neurodegenerative Diseases

### 2019-03-08 paper/paper.tex
minor

\begin{abstract}
We introduce Disease Knowledge Transfer (DKT), a novel technique for transferring biomarker information between related neurodegenerative diseases. DKT infers robust multimodal biomarker trajectories in rare neurodegenerative diseases even when only limited, unimodal data is available, by transferring information from larger multimodal datasets from common neurodegenerative diseases. DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases. As opposed to current deep learning approaches, DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects instead of solving simpler control vs diseased classification tasks. Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects. For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from the Dementia Research Centre (DRC) UK, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available. We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature. We further validate DKT in two situations: (1) on synthetic data, showing that it can accurately estimate the ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to standard approaches. While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases. Source code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.

@@

We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal biomarker progressions for multiple diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes. This is achieved by exploiting biomarker relationships that are shared across diseases, whilst accounting for differences in the spatial distribution of brain pathology. DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects at risk of diseases. We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data. This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from the Dementia Research Centre (DRC), UK. We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings. We further validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients, showing it yields favourable performance compared to standard approaches. Code for DKT is available online: \url{https://github.com/mrazvan22/dkt}.

\begin{figure}[h]
 \centering
 \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{disease_knowledge_transfer_symbols.pdf}
 \caption{Diagram of the proposed DKT framework. We assume that each disease can be modelled as the evolution of abstract dysfunction scores (Y-axis, top row), each one related to different brain regions. Each region-specific dysfunction score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region. For instance, the temporal dysfunction, modelled as a biomarker in the disease specific model (top row), is the X-axis in the disease agnostic model (temporal unit, bottom row), which aggregates together abnormality from amyloid, tau and MR imaging within the temporal lobe. The biomarker relationships within the bottom units are assumed to be disease agnostic and shared across all diseases modelled. Disease knowledge transfer can then be achieved via the disease-agnostic units. Mathematical notation from section \ref{sec:method} is shown in red to ease understanding.}
 \label{fig:diagram}
\end{figure}

@@

\subsection{Disease-Specific Model}

@@

\subsection{Disease-Agnostic Model}

@@

\subsection{Extending to Multiple Subjects and Biomarkers}

@@

% Fig1: biomarker traj. over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
Results on TADPOLE and DRC datasets are presented in Fig. \ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data. The model shows an unbiased data fit (Fig. \ref{fig:pcaTadDisSpace}A), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunction scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}. We also show the progression of dysfunction scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C). While both typical AD and PCA show early hippocampal dysfunction, PCA shows higher dysfunction in the occipital, temporal and parietal regions in late stages, while tAD shows widespread dysfunction, in line with previous literature findings \cite{crutch2012posterior}.

>>>>

\begin{abstract}
We introduce Disease Knowledge Transfer (DKT), a novel technique for transferring biomarker information between related neurodegenerative diseases. DKT infers robust multimodal biomarker trajectories in rare neurodegenerative diseases even when only limited, unimodal data is available, by transferring information from larger multimodal datasets from common neurodegenerative diseases. DKT is a joint-disease generative model of biomarker progressions, which exploits biomarker relationships that are shared across diseases. As opposed to current deep learning approaches, DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects instead of solving simpler control vs diseased classification tasks. Here we demonstrate DKT on Alzheimer's disease (AD) variants and its ability to predict trajectories for multimodal biomarkers in Posterior Cortical Atrophy (PCA), in lack of such data from PCA subjects. For this we train DKT on a combined dataset containing subjects with two distinct diseases and sizes of data available: 1) a larger, multimodal typical AD (tAD) dataset from the TADPOLE Challenge, and 2) a smaller unimodal Posterior Cortical Atrophy (PCA) dataset from the Dementia Research Centre (DRC) UK, for which only a limited number of Magnetic Resonance Imaging (MRI) scans are available. We first show that the estimated multimodal trajectories in PCA are plausible as they agree with previous literature. We further validate DKT in two situations: (1) on synthetic data, showing that it can accurately estimate the ground truth parameters and (2) on 20 DTI scans from controls and PCA patients, showing that it has favourable predictive performance compared to standard approaches. While we demonstrated DKT on Alzheimer's variants, we note DKT is generalisable to other forms of related neurodegenerative diseases.

@@

We propose Disease Knowledge Transfer (DKT), a generative joint model that estimates continuous multimodal biomarker progressions for multiple diseases simultaneously -- including rare neurodegenerative diseases -- and which inherently performs transfer learning between the modelled phenotypes. This is achieved by exploiting biomarker relationships that are shared across diseases, whilst accounting for differences in the spatial distribution of brain pathology. DKT is interpretable, which allows us to understand underlying disease mechanisms, and can also predict the future evolution of subjects at risk of diseases. We apply DKT on Alzheimer's variants and demonstrate its ability to predict non-MRI trajectories for patients with Posterior Cortical Atrophy, in lack of such data. This is done by fitting DKT to two datasets simultaneously: (1) the TADPOLE Challenge \cite{marinescu2018tadpole} dataset containing subjects from the Alzheimer's Disease Neuroimaging Initiative (ADNI) with MRI, FDG-PET, DTI, AV45 and AV1451 scans and (2) MRI scans from patients with Posterior Cortical Atrophy from the Dementia Research Centre (DRC), UK. We first show that the estimated trajectories for PCA and tAD subjects are plausible as they agree with previous literature findings. We finally validate DKT on simulated data from two synthetic diseases with known ground truth, and a set of 20 DTI scans from controls and PCA patients.

\begin{figure}[h]
 \centering
 \includegraphics[width=0.9\textwidth,trim=0 0 0 0,clip]{disease_knowledge_transfer_symbols.pdf}
 \caption{Diagram of the proposed DKT framework. We assume that each disease can be modelled as the evolution of abstract dysfunction scores (Y-axis, top row), each one related to different brain regions. Each region-specific dysfunction score then further models (X-axis, bottom row) the progression of several modality-specific biomarkers within that same region. For instance, the temporal dysfunction, modelled as a biomarker in the disease specific model (top row), is the X-axis in the disease agnostic model (temporal unit, bottom row), which aggregates together abnormality from amyloid, tau and MR imaging within the temporal lobe. The biomarker relationships within the bottom units are assumed to be disease agnostic and shared across all diseases modelled. Knowledge transfer between the two diseases can then be achieved via the disease-agnostic units. Mathematical notation from section \ref{sec:method} is shown in red to ease understanding.}
 \label{fig:diagram}
\end{figure}

@@

% \subsection{Disease-Specific Model}

@@

% \subsection{Disease-Agnostic Model}

@@

% \subsection{Extending to Multiple Subjects and Biomarkers}

@@

% Fig1: biomarker traj. over dysfunction scores in one agnostic unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
Results on TADPOLE and DRC datasets are presented in Fig. \ref{fig:pcaTadDisSpace}, showing the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with aligned subject data. The population model shows a good data fit (Fig. \ref{fig:pcaTadDisSpace}A, MAE $\leq$ 0.25), and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunction scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}. We also show the progression of dysfunction scores over the disease stage for typical AD (Fig \ref{fig:pcaTadDisSpace}B) and PCA (Fig \ref{fig:pcaTadDisSpace}C). While both typical AD and PCA show early hippocampal dysfunction, PCA shows higher dysfunction in the occipital, temporal and parietal regions in late stages, while tAD shows widespread dysfunction, in line with previous literature findings \cite{crutch2012posterior}.
